---
title: "Morphology x Gene Expression DIABLO (moa-fine)"
output: html_notebook
---

This implements mixOmics DIABLO model (`block.splsda`) that integrates the
HepG2 morphology and gene expression blocks and tries to predict **`moa-fine`** compound annotation
stored in `data/hepg2_mdata.h5mu`.

The X blocks (gex, morph) are read from the wide CSVs generated from the mudata.
The `moa-fine` labels aren't in those CSVs, so they were exported to `data/mixOmics/hepg2_moa_labels.csv`

```{r export-note, eval=FALSE}
# One-off export (Python, "hackathon" conda env) — re-run only if hepg2_mdata.h5mu changes:
#   import mudata as md
#   mdata = md.read_h5mu("data/hepg2_mdata.h5mu")
#   obs = mdata.obs[["drug", "moa-fine"]].reset_index(drop=True)
#   obs.columns = ["drug", "moa_fine"]
#   obs.to_csv("data/mixOmics/hepg2_moa_labels.csv", index=False)
```

The gene pre-filter in step 4 reuses the same highly-variable-gene call
`wl_integration_scvi.ipynb` uses for its scVI model (`seurat_v3` flavor on
raw counts), rather than re-deriving "top variable" genes independently in
R, so the gex feature set is identical across the two notebooks. Also a
one-off export, re-run only if `hepg2_mdata.h5mu` changes:

```{r export-note-hvg, eval=FALSE}
#   import mudata as md
#   import scanpy as sc
#   mdata = md.read_h5mu("data/hepg2_mdata.h5mu")
#   gex = mdata["gex"]
#   sc.pp.highly_variable_genes(gex, flavor="seurat_v3", n_top_genes=5000, layer="raw_counts")
#   hvg = gex.var[["highly_variable"]].reset_index(names="gene")
#   hvg.to_csv("data/mixOmics/gex_hvg_scvi.csv", index=False)
```

```{r setup}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("mixOmics", quietly = TRUE)) BiocManager::install("mixOmics")
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
```

```{r libraries, message=FALSE}
library(mixOmics)
library(tidyverse)
setwd("~/Git/hackathon_crossmodal_stream/OpenScreen/")
```

## 1. Load and reshape the X blocks

```{r load-data}
data_dir <- "~/Git/hackathon_crossmodal_stream/OpenScreen/data/mixOmics"

gex_raw <- read_csv(file.path(data_dir, "hepg2_exp_X.csv"), show_col_types = FALSE)
morph_raw <- read_csv(file.path(data_dir, "hepg2_morph_X.csv"), show_col_types = FALSE)

# Both files are wide (feature x compound): first column holds feature names,
# remaining columns are compounds. mixOmics wants compounds (samples) in rows
# and features in columns, so transpose.
to_sample_by_feature <- function(df) {
  feat_names <- df[[1]]
  mat <- as.matrix(df[, -1])
  storage.mode(mat) <- "numeric"
  rownames(mat) <- feat_names
  t(mat)
}

gex <- to_sample_by_feature(gex_raw)
morph <- to_sample_by_feature(morph_raw)

dim(gex)   # compounds x genes
dim(morph) # compounds x morphology features
```

## 2. Load moa-fine labels and align everything

```{r load-labels}
labels <- read_csv(file.path(data_dir, "hepg2_moa_labels.csv"), show_col_types = FALSE)

common_compounds <- Reduce(intersect, list(rownames(gex), rownames(morph), labels$drug))
length(common_compounds)

gex <- gex[common_compounds, , drop = FALSE]
morph <- morph[common_compounds, , drop = FALSE]
labels <- labels[match(common_compounds, labels$drug), ]

stopifnot(identical(rownames(gex), rownames(morph)))
stopifnot(identical(rownames(gex), labels$drug))
```

## 3. Clean up the outcome (`moa-fine`)

`moa-fine` has ~20 classes over 119 compounds, more than half of which are
`"unclear"`, and many real classes have only 1-2 members — too few for any
cross-validated fold. `"unclear"` is **excluded from training**
but the compounds are kept aside so we can later project them onto the fitted
embedding with `predict()` and see where they land relative to the trained
classes. Classes too small to survive `Mfold` CV are dropped.

```{r clean-labels}
min_class_n <- 3 # should be >= the number of CV folds used below

y_all <- factor(labels$moa_fine)
table(y_all)

is_unclear <- y_all == "unclear"
unclear_compounds <- rownames(gex)[is_unclear]
length(unclear_compounds)

class_counts <- table(y_all[!is_unclear])
small_classes <- names(class_counts[class_counts < min_class_n])

train_keep <- !is_unclear & !(y_all %in% small_classes)
Y <- droplevels(y_all[train_keep])
train_compounds <- rownames(gex)[train_keep]

table(Y)
length(Y)
```

```{r}
print("MoA classes too small for training (fewer than min_class_n)")
cat(small_classes, sep = '\n')
```

## 4. Quality checks and gene pre-filter

The zero-variance filter is computed from the **training** compounds only,
so the held-out `unclear` compounds don't leak into which features get
kept. The top-gene filter, though, reuses the exact highly-variable-gene
set `wl_integration_scvi.ipynb` selects for its scVI model (`seurat_v3` on
raw counts, `n_top_genes=5000`) rather than a fresh in-R variance ranking —
that selection was fit on all 119 compounds (raw counts, not the z-scored
values used here), so unlike the zero-variance filter it isn't
training-only, but it does keep the two notebooks' gex feature sets
identical for a fair comparison.

```{r qc}
gex_train <- gex[train_compounds, , drop = FALSE]
morph_train <- morph[train_compounds, , drop = FALSE]

sum(is.na(gex_train))
sum(is.na(morph_train))

keep_gex_cols <- apply(gex_train, 2, var, na.rm = TRUE) > 0
keep_morph_cols <- apply(morph_train, 2, var, na.rm = TRUE) > 0
gex_train <- gex_train[, keep_gex_cols, drop = FALSE]
morph_train <- morph_train[, keep_morph_cols, drop = FALSE]

# Same 5000 genes as wl_integration_scvi.ipynb's scVI model (see the
# export-note-hvg chunk above), intersected with the zero-variance filter
# above -- a gene can be HVG-flagged genome-wide but still constant across
# just these 47 training compounds.
hvg <- read_csv(file.path(data_dir, "gex_hvg_scvi.csv"), show_col_types = FALSE)
hvg_genes <- hvg$gene[hvg$highly_variable]
hvg_genes_kept <- intersect(hvg_genes, colnames(gex_train))
length(hvg_genes_kept) # vs. length(hvg_genes) if some got dropped as zero-variance here

gex_train <- gex_train[, hvg_genes_kept, drop = FALSE]

dim(gex_train)
dim(morph_train)
```

```{r unclear-features}
# Same feature set as the training blocks above, applied to the held-out
# 'unclear' compounds, for projection later on.
gex_unclear <- gex[unclear_compounds, colnames(gex_train), drop = FALSE]
morph_unclear <- morph[unclear_compounds, colnames(morph_train), drop = FALSE]

dim(gex_unclear)
dim(morph_unclear)
```

```{r rename-train}
# Keep downstream chunks (design/model-fitting) unchanged: from here on
# `gex`/`morph` refer to the training blocks only.
gex <- gex_train
morph <- morph_train
```

## 5. Assemble blocks and design matrix

```{r data-design}
data <- list(
  gex = gex,
  morph = morph
)

# Connection strength between the X blocks. Small values favour prediction,
# values close to 1 favour exploring shared/correlated structure between
# blocks. block.splsda internally adds Y and links every block to it.
design <- matrix(0.8, ncol = length(data), nrow = length(data),
                  dimnames = list(names(data), names(data)))
diag(design) <- 0
design
```

## 6. Pick the number of components

```{r ncomp-tuning}
# near.zero.var = TRUE: with 10-fold CV and ~47 training compounds, a fold's
# ~42-sample training subset can easily have a gene/feature that comes out
# constant even though it varies across the full training set. Without this,
# perf() below errors with "There are features with zero variance in block
# 'gex'." the moment a fold+repeat combination hits that. block.plsda applies
# nearZeroVar() up front and perf() re-applies the same filter inside every
# CV refit, so it's fixed for the whole CV run, not just the first fit.
diablo.init <- block.plsda(X = data, Y = Y, ncomp = 10, design = design, near.zero.var = TRUE)

set.seed(123)
perf.diablo <- perf(diablo.init, validation = "Mfold", folds = 3, nrepeat = 10)

plot(perf.diablo)
perf.diablo$choice.ncomp$WeightedVote
```


```{r choose-ncomp}
ncomp <- perf.diablo$choice.ncomp$WeightedVote["Overall.BER", "max.dist"]
ncomp <- max(2, ncomp) # floor of 2 so comp1 vs comp2 plots below have something to show
ncomp
```


## 7. Tune sparsity (feature selection) per block/component

```{r keepx-grid}
test.keepX <- list(
  gex = c(5, 10, 20, 30, 50, 75, 100),
  morph = c(5, 10, 20, 30, 50)
)

set.seed(123)
tune.diablo <- tune.block.splsda(
  X = data, Y = Y, ncomp = ncomp,
  test.keepX = test.keepX, design = design,
  validation = "Mfold", folds = 3, nrepeat = 5,
  dist = "max.dist", progressBar = TRUE,
  near.zero.var = TRUE
)

tune.diablo$choice.keepX
```

## 8. Fit the final DIABLO model

```{r final-model}
diablo <- block.splsda(
  X = data, Y = Y, ncomp = ncomp,
  keepX = tune.diablo$choice.keepX,
  design = design,
  near.zero.var = TRUE
)

diablo
```

## 8b. Single-modality DIABLO baselines (gex-only, morph-only)

To find out whether the joint model is actually using cross-modal structure,
fit the *same* `block.splsda` machinery on each block alone, reusing the
joint model's `ncomp` and each block's own tuned `keepX` — the only thing
that changes is whether the other block is present, so the three embeddings
(joint, gex-only, morph-only) are directly comparable rather than being
different methods with independently-chosen hyperparameters. `block.splsda`
accepts a length-1 `X` list without complaint (the design matrix just
collapses to a 1x1 zero matrix, since there are no other blocks to connect).

```{r single-block-models}
design.gex <- matrix(0, 1, 1, dimnames = list("gex", "gex"))
diablo.gex <- block.splsda(
  X = list(gex = data$gex), Y = Y, ncomp = ncomp,
  keepX = list(gex = tune.diablo$choice.keepX$gex),
  design = design.gex,
  near.zero.var = TRUE
)

design.morph <- matrix(0, 1, 1, dimnames = list("morph", "morph"))
diablo.morph <- block.splsda(
  X = list(morph = data$morph), Y = Y, ncomp = ncomp,
  keepX = list(morph = tune.diablo$choice.keepX$morph),
  design = design.morph,
  near.zero.var = TRUE
)

diablo.gex
diablo.morph
```

## 9. Evaluate

```{r perf-final}
set.seed(123)
perf.final <- perf(diablo, validation = "Mfold", folds = 3, nrepeat = 10,
                     dist = "max.dist", progressBar = FALSE)

perf.final$WeightedVote.error.rate
```

## 10. Visualise

```{r plot-diablo}
plotDiablo(diablo, ncomp = 1)
```

```{r plot-indiv}
plotIndiv(diablo, ind.names = FALSE, legend = TRUE,
          title = "DIABLO: sample space per block")
```

```{r plot-var}
plotVar(diablo, style = "graphics", legend = TRUE,
        pch = c(16, 17), cex = c(2, 2),
        title = "DIABLO correlation circle")
```

```{r plot-loadings}
pdf(')
plotLoadings(diablo, comp = 1, contrib = "max", method = "median")
```

```{r circos}
circosPlot(diablo, cutoff = 0.7, line = TRUE)
```

```{r cim}
cimDiablo(diablo)
```

```{r network}
ntw <- network(diablo, blocks = c("gex", "morph"), cutoff = 0.6,
                color.node = c("steelblue", "darkorange"))
```

## 11. Export selected features per block/component

```{r selected-features}
selected <- map(seq_len(ncomp), function(cp) {
  map_dfr(names(data), function(bl) {
    sv <- selectVar(diablo, block = bl, comp = cp)[[bl]]
    tibble(component = cp, block = bl, feature = sv$name, loading = sv$value$value.var)
  })
})
selected <- bind_rows(selected)

selected

write_csv(selected, file.path(data_dir, "diablo_moa_selected_features.csv"))
```

## 12. Project the excluded `unclear` compounds onto each embedding

`unclear` was excluded from training in step 3, but `predict()` can still map
those compounds' gex/morph profiles into an already-fitted DIABLO latent
space, using that model's own loadings. This gives their component scores
(and, for curiosity, which trained class they'd be nearest to) without them
ever having influenced any of the three models fit above.

```{r predict-unclear}
pred.unclear.joint <- predict(diablo, newdata = list(gex = gex_unclear, morph = morph_unclear))
pred.unclear.gex <- predict(diablo.gex, newdata = list(gex = gex_unclear))
pred.unclear.morph <- predict(diablo.morph, newdata = list(morph = morph_unclear))

# Nearest trained class per block/component, per distance metric (curiosity
# only -- these compounds were never labelled with a real moa-fine class)
pred.unclear.joint$class$max.dist
```

```{r plot-projection}
plot_df <- map_dfr(names(data), function(bl) {
  train_df <- as.data.frame(diablo$variates[[bl]][, 1:2, drop = FALSE])
  colnames(train_df) <- c("comp1", "comp2")
  train_df$compound <- rownames(train_df)
  train_df$moa_fine <- as.character(Y)
  train_df$source <- "training"

  new_df <- as.data.frame(pred.unclear.joint$variates[[bl]][, 1:2, drop = FALSE])
  colnames(new_df) <- c("comp1", "comp2")
  new_df$compound <- rownames(new_df)
  new_df$moa_fine <- "unclear"
  new_df$source <- "projected (unclear)"

  bind_rows(train_df, new_df) %>% mutate(block = bl)
})

ggplot(plot_df, aes(comp1, comp2, color = moa_fine, shape = source)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_shape_manual(values = c(training = 16, `projected (unclear)` = 4)) +
  facet_wrap(~block, scales = "free") +
  theme_bw() +
  labs(title = "DIABLO embedding with projected 'unclear' compounds",
       x = "Component 1", y = "Component 2")
```

## 13. Export all three embeddings for cross-tool (scanpy) comparison

The gex-only and morph-only models each have a single block, so their
`$variates` for that block *is* the embedding. The joint model instead has
one variate matrix per block (correlated, but not identical) — the standard
single embedding to report for DIABLO is the per-component weighted average
of the block variates, weighted by each block's correlation with `Y`
(`diablo$weights`; this is exactly what `plotIndiv(diablo, blocks =
"weighted.average")` plots). Computed here directly, applied to both the
training variates and the `predict()`-projected ones, so it plays the same
role as totalVI's single joint latent (`X_totalvi_joint`) in
`wl_integration_scvi.ipynb`.

```{r joint-embedding}
weighted_average_variates <- function(variates, weights, blocks) {
  w <- weights[blocks, , drop = FALSE]
  weighted <- Map(function(bl) sweep(variates[[bl]], 2, as.numeric(w[bl, ]), `*`), blocks)
  sweep(Reduce(`+`, weighted), 2, colSums(w), `/`)
}

diablo.joint.train.variates <- weighted_average_variates(diablo$variates, diablo$weights, names(data))
diablo.joint.pred.variates <- weighted_average_variates(pred.unclear.joint$variates, diablo$weights, names(data))
```

One CSV per embedding (joint / gex-only / morph-only), each with all 119
compounds: the `ncomp` DIABLO component scores (fitted for the 64 training
compounds, `predict()`-projected for the 55 `unclear` ones), plus `moa_fine`
and whether that row was `training` or `projected`.

```{r export-embeddings}
comp_cols <- paste0("comp", seq_len(ncomp))

variates_to_df <- function(train_variates, pred_variates, train_Y) {
  train_df <- as.data.frame(train_variates[, seq_len(ncomp), drop = FALSE])
  colnames(train_df) <- comp_cols
  train_df$compound <- rownames(train_variates)
  train_df$moa_fine <- as.character(train_Y)
  train_df$source <- "training"

  new_df <- as.data.frame(pred_variates[, seq_len(ncomp), drop = FALSE])
  colnames(new_df) <- comp_cols
  new_df$compound <- rownames(pred_variates)
  new_df$moa_fine <- "unclear"
  new_df$source <- "projected"

  bind_rows(train_df, new_df) %>% relocate(compound, moa_fine, source)
}

embeddings <- list(
  joint = variates_to_df(diablo.joint.train.variates, diablo.joint.pred.variates, Y),
  gex   = variates_to_df(diablo.gex$variates$gex, pred.unclear.gex$variates$gex, Y),
  morph = variates_to_df(diablo.morph$variates$morph, pred.unclear.morph$variates$morph, Y)
)

embed_dir <- file.path(data_dir, "diablo_embeddings")
dir.create(embed_dir, showWarnings = FALSE)

iwalk(embeddings, function(df, name) {
  write_csv(df, file.path(embed_dir, paste0("diablo_", name, ".csv")))
})

map_int(embeddings, nrow) # sanity check: 119 rows each
```
